Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela

### 2024-12-18 agujournaltemplate.tex
presync

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all depth discrete measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

@@

    %RUE/BDEF ANALYSIS
    To investigate the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity, and resource use efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index, and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling. $RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
    We excluded observations with $NO_3$ concentrations below 0.01 µM (n = 36) because they are close to the detection limit, resulting in imprecise measurements \cite{ptacnik_diversity_2008}.
    To quantify the trend in the relationships, we use linear regression models using the "lm" function \cite{r_core_team_r_2024}.
    %here I want to extend to glmm that correct for depth and seasonality for the final results

@@

% Figure 3
The effects of diversity on ecosystem functions can be described by the relationship between measured diversity indicators and corresponding measurements of biomass, productivity, and resource use efficiency (RUE). In order to take into account effects of depth on the resulting response, we used the same interpolated depth intervals (0-25, 25-50, 50-75, and 75-100 meters) and plotted coincidental measurements against each other. The considered diversity indices are the genus richness, Shannon index, and Pielou index, which are plotted against chlorophyll a biomass, primary productivity, as well as RUE for Nitrate ($RUE_N$). Immediately apparent, is that depth has a strong effect on the measured relationship, in particular for genus richness, where most instances of high biomass, productivity and $RUE_N$ at high diversity were observed in the uppwer water column (see Figure \ref{fig:bdef} $a, d, g$). For the Shannon index, diversity at the different depths shows a much wider variability and overlap, so the pattern is less clear.  Similarly, community evenness shows no clear depth structure, except that very high biomass and productivity measurements in the upper 25 meters tend to be clustered at intermediate evenness. 
To quantify the relationships, we performed simple linear regression... %TODO perhaps check the glmm method to account for depth and seasonality

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlayed with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards postive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

% Figure 5
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 5 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower. The 21 \degree C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community. The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data is biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms. However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the paramater range. The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group. For the AMO index, we see a clear pattern that negative anomalies highly affect the community. This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes. Nitrate concentrations between 6 and 8 $µM$ is where the community changed most, which indicates these values as a threshold for the composition of the community. For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e. high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \degree C) and high salinity (above 36.8 PSU). One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.

@@

\begin{figure}
\label{fig:bdef}
\noindent\includegraphics[width=\textwidth]{fig/BDEF_depth_PLOT.pdf}
\caption{Linear regression of diversity indices to log-transformed Chlorophyll Biomass, Primary Production, Resource Use Efficiency for nitrate ($RUE_N$) and for phosphate ($RUE_P$). $RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively). All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\textquotesingle s Evenness Index.}
\end{figure}

@@

\begin{figure}
\label{fig:GF}
\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3.pdf}
\caption{Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm. What I also included in the plot above is the underlying mechanism of fitting a random forest model for each genus. The plot on the top right shows the weighted importance for each genus, ranked according to the predictive capability of the model, and color coded by their respective group. Similarly, the cumulative importance plots below show the overall importance, as well as the individual color lines that show the cumulative importance per genus predicted. This is not necessarily useful to show in the final manuscript, unless we want to discuss the model output down to the genus level. Not surprisingly, it is mostly diatoms that are predictable, but also other groups are present here.}
\end{figure}

>>>>

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

@@

@@

% Figure 3
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlayed with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards postive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

% Figure 4
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 5 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower. The 21 \degree C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community. The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data is biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms. However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the paramater range. The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group. For the AMO index, we see a clear pattern that negative anomalies highly affect the community. This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes. Nitrate concentrations between 6 and 8 $µM$ is where the community changed most, which indicates these values as a threshold for the composition of the community. For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e. high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \degree C) and high salinity (above 36.8 PSU). One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.

@@

@@

\begin{figure}
\label{fig:GF}
\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
\caption{Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm. What I also included in the plot above is the underlying mechanism of fitting a random forest model for each genus. The plot on the top right shows the weighted importance for each genus, ranked according to the predictive capability of the model, and color coded by their respective group. Similarly, the cumulative importance plots below show the overall importance, as well as the individual color lines that show the cumulative importance per genus predicted. This is not necessarily useful to show in the final manuscript, unless we want to discuss the model output down to the genus level. Not surprisingly, it is mostly diatoms that are predictable, but also other groups are present here.}
\end{figure}
